visualFields 1.0.7: Tools for Analyzing the Field of Vision in \proglang{R}

### 2025-05-17 jss-article-rnw/article.Rnw
Fixed problem with merge: ganglion cell displacement plot added.
Cleaned up the manuscript.

%% another package (only for this demo article)
\usepackage{framed}

@@

<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("MASS")
library("plotrix")
library("ggplot2")
@

@@

\begin{document}
%%\SweaveOpts{concordance=TRUE}

@@

@@

In healthy vision, most visual functions, particularly light sensitivity, decline with increasing eccentricity—i.e., distance from the center of fixation. This decline in differential light sensitivity forms a three-dimensional structure known as the hill of vision, also referred to as Traquair’s island of vision \citep{IMF:Traquair1938}. Figure~\ref{hill} below shows a classic rendering of the hill of vision by Aulhorn and Harms \citep{IMF:Aul1967}.

@@

At the conclusion of the test, the inverse of the contrast threshold, $\Delta L_{\text{threshold}}$, is recorded at each location using a logarithmic decibel (dB) scale. This scale is defined as the attenuation from the maximum possible stimulus contrast, $\Delta L_{\text{max}}$:

@@

For common devices such as the Humphrey Field Analyzer and the Octopus 900, the maximum luminance differential is $\Delta L_{\text{max}} = \frac{10{,}000\ \text{apostilbs}}{\pi} \approx 3183.1\ \text{cd},\text{m}^{-2}$.

@@

\begin{enumeration}
  \item [False negatives (FNR)] Missed responses to stimuli that were previously seen.
  \item [False positives (FPR)] Responses when no stimulus is presented
  \item [Fixation losses] Responses to stimuli presented within the blind spot
\end{enumeration}

@@

Test locations are defined by location maps (\texttt{locmap}), which specify the spatial coordinates of stimulus presentation points in degrees of visual angle. These maps reflect the layout used by visual field testing devices. When analyzing data, the correct \texttt{locmap} must be specified.

By default, the \texttt{24-2} pattern is used, consisting of 54 regularly spaced locations across the central visual field.

<<>>=
library(visualFields)
default_locmap <- getlocmap()
str(default_locmap)
@

Each \texttt{locmap} contains:
\begin{itemize}
  \item A coordinate table with $x$- and $y$-positions (in degrees),
  \item A vector \texttt{bs} indicating locations corresponding to the physiological blind spot.
\end{itemize}

Other patterns are available in the package, including the \texttt{10-2} grid:

<<>>=
ten_two_locmap <- locmaps$p10d2
str(ten_two_locmap)
@

All available patterns can be listed with:

@@

If the blind spot is outside the tested region (e.g., in the 10$^\circ$ pattern) or is skipped (e.g., in the Octopus G1), the \texttt{bs} field is set to \texttt{numeric(0)}.

\subsection{Loading data}

The package includes example data from a glaucoma patient (\texttt{vfpwgSunyiu24d2}), consisting of 42 visual field tests.

<<>>=
str(vfpwgSunyiu24d2[, 1:10])
@

Each row represents a monocular test. The eye is labeled using Latin abbreviations: OD (right eye), OS (left eye), and OU (both eyes). Initial columns store patient information and reliability metrics. Actual test values begin at the column number stored in the \texttt{locini} variable:

<<>>=
getlocini()
getvfcols()
vfpwgSunyiu24d2[1, getvfcols()]
@

No columns should appear after these tsest values. Use \texttt{vffilter} to select specific tests and \texttt{vfisvalid} to verify data structure:

@@

\subsection{Statistical analysis of contrast thresholds}

\subsubsection{Defining defective locations}

@@

To detect abnormal points, we compare observed sensitivities with an age-corrected mean normal field. These comparisons require a \textbf{normative dataset} and location-wise \textbf{age-linear models}.

The difference between observed and normal values is the \textbf{total deviation (TD)}. In this package, negative values indicate lower-than-normal sensitivity, following the Humphrey Field Analyzer convention. For perimeters that reverse this definition (e.g., Octopus), users must invert the sign.

To reveal focal defects masked by generalized depression (e.g., due to cataract), a \textbf{pattern deviation (PD)} is computed by subtracting a general height (GH) estimate from the TD. For 24-2, GH is defined as the 7th highest TD value.

\subsubsection{Normative datasets}

The package includes three normative datasets:

\begin{itemize}
  \item SUNY-IU dataset for 24-2 with Goldmann size III \citep{IMF:Mar2013b}
  \item Iowa PC26 and Peri datasets for Goldmann size V \citep{IMF:Marin2018c}
\end{itemize}

<<>>=
str(vfctrSunyiu24d2[, 1:10])
str(vfctrIowaPC26[, 1:10])
str(vfctrIowaPeri[, 1:10])
@

To build the normative model:

@@

The normative object \texttt{nv} includes coefficients of age-linear models:

<<>>=
print(nv$agem$coeff)
@

The expected field at age 50 is:

<<>>=
age <- 50
print(nv$agem$model(age))
@

Blind spot locations are returned as \texttt{NA}.

@@

Total deviation is the difference between observed and expected sensitivity:

@@

Or, use the automated function:

@@

Global indices summarize the visual field:

<<>>=
gl <- getgl(example2)
print(gl[, -(1:getlocini() - 1)])
@

\begin{itemize}
  \item \texttt{msens}: Mean sensitivity (MS)
  \item \texttt{ssens}: SD of sensitivity
  \item \texttt{tmd}: Mean total deviation (MD)
  \item \texttt{tsd}: SD of total deviation
  \item \texttt{pmd}: Mean pattern deviation
  \item \texttt{psd}: SD of pattern deviation
  \item \texttt{gh}: General height
  \item \texttt{vfi}: Visual field index (VFI) \citep{IMF:Ben2008}
\end{itemize}

Associated probabilities:

<<>>=
print(getglp(gl)[, -(1:getlocini() - 1)])
@

@@

Standard displays include color-coded probability maps, graphical representations of the blind spot, and border enhancements that encode diagnostic information. For example, 	exttt{vfplot} with the argument 	exttt{"td"} generates plots that require normative data on differential light sensitivity and the probabilities of deviations from these norms. This information is included for the Humphrey Field Analyzer patterns but must be provided separately for other perimeters.

@@

It is important to specify the type of data represented in the visual field tables. The default is set to \textbf{total deviation} (	exttt{td}).

@@

Progression analysis methods fall into two broad categories: \textbf{event-based} and \textbf{trend-based} approaches. These may further be classified by their reliance on population-based normative data, mathematical models of progression, or model-free methods.

@@

\subsection{Plotting progression}

@@

\subsection{Statistical analysis}

Statistical progression analysis may be performed via:
\begin{itemize}
\item \textbf{Global linear regression} (e.g., on MD)
\item \textbf{Pointwise linear regression} (PLR)
\end{itemize}

The \texttt{glr()} function estimates a regression model for global indices. For the example series, the patient loses \texttt{getgl(example3)} dB/year, with $p < \texttt{getgl(example3)}$.

%\texttt{r round(abs(glr(getgl(example3))$sl), 2)}
%\texttt{r format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)}

Pointwise linear regression is performed using \texttt{plr()}, resulting in one slope and $p$-value per location:

<<>>=
prog1 <- plr(example3)
str(prog1, max.level = 1)
print(sum(prog1$pval[-getlocmap()$bs] < 0.05))
@

@@

PLR-based criteria (e.g., 3 or more progressing points) are clinically useful but lack explicit control for specificity. \textbf{Permutation of Pointwise Linear Regression (PoPLR)} addresses this by combining the 52 $p$-values into a single $S$-statistic (per Fisher's method) and comparing it to a null distribution from permuted series [@IMF:Ole2012; @IMF:Marin2021].

<<>>=
prog2 <- poplr(example3)
str(prog2, max.level = 1)
@

@@

\subsection{Single field reports}

@@

The ganglion cell bodies stimulated by a visual field location are displaced from the photoreceptor input region, especially in the central field. This displacement is modeled in \pkg{visualFields} based on Montesano et al. [@IMF:Montesano2020].

@@

Nerve fiber paths from each location to the optic disc can also be modeled, using Jansonius' framework [@IMF:Jan2009; @IMF:Jan2012].

<<>>=
psi23 <- loc2psi(vf2gc(locmaps$p24d2$coord[23,]))
ang23 <- psi2oct(psi23)
print(ang23 - 360)
@

\subsection{Progression reports}

@@

Perimetry conventions vary across manufacturers. The package includes normative data for Humphrey 24-2 (Goldmann III) and Iowa full-field perimetry with Goldmann V [@IMF:Marin2018c]. The \texttt{loadoctopus} function provides partial support for Octopus G1 fields, but normative support is incomplete.

Users may provide custom normative datasets using \texttt{nvgenerate()} and \texttt{setnv()}.

@@

Each analysis requires environment variables:
\begin{itemize}
\item \texttt{nv}: normative values
\item \texttt{locmap}: test grid
\item \texttt{gpar}: graphical parameters
\item \texttt{locini}: first sensitivity column
\end{itemize}

<<>>=
print(names(locmaps))
print(names(normvals))
print(names(gpars))
@

@@

Custom test structures must:
\begin{itemize}
\item Start with columns: \texttt{id}, \texttt{eye}, \texttt{date}, \texttt{time}
\item Include \texttt{age}, optionally \texttt{fpr}, \texttt{fnr}, \texttt{fl}
\item Have sensitivity columns named \texttt{l1}--\texttt{lk}
\end{itemize}

@@

If normative data exist, use \texttt{vfread()}, \texttt{nvgenerate()}, and \texttt{setnv()} to enable deviation computations.

>>>>

%% another package (only for this demo article)
\usepackage{framed}
\usepackage{listings}

@@

<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", width = 60, useFancyQuotes = FALSE)
library("MASS")
library("plotrix")
library("ggplot2")
library("utils")
library("visualFields")
@

@@

\begin{document}
%%\SweaveOpts{concordance=TRUE}
\lstset{breaklines=true, breakatwhitespace=true}

@@

\section[Introduction: Visual Fields in R]{Introduction: Visual Fields} \label{sec:intro}

@@

In healthy vision, most visual functions, particularly light sensitivity, decline with increasing eccentricity—i.e., distance from the center of fixation. This decline in differential light sensitivity forms a three-dimensional structure known as the hill of vision, also referred to as Traquair’s island of vision \citep{IMF:Traquair1938}. Figure~\ref{fig:hill} below shows a classic rendering of the hill of vision by Aulhorn and Harms \citep{IMF:Aul1967}.

@@

At the conclusion of the test, the inverse of the contrast threshold, $\Delta L_{threshold}$, is recorded at each location using a logarithmic decibel (dB) scale. This scale is defined as the attenuation from the maximum possible stimulus contrast, $\Delta L_{max}$:

@@

For common devices such as the Humphrey Field Analyzer and the Octopus 900, the maximum luminance differential is $\Delta L_{max} = \frac{10{,}000\ apostilbs}{\pi} \approx 3183.1\ cd,{m}^{-2}$.

@@

\begin{description}
  \item [False negatives (FNR)] Missed responses to stimuli that were previously seen.
  \item [False positives (FPR)] Responses when no stimulus is presented
  \item [Fixation losses] Responses to stimuli presented within the blind spot
\end{description}

@@

Test locations are defined by location maps (\code{locmaps}), which specify the spatial coordinates of stimulus presentation points in degrees of visual angle. These maps reflect the layout used by visual field testing devices. When analyzing data, the correct \code{locmap} must be specified.

@@

By default, the \code{24-2} pattern is used, consisting of 54 regularly spaced locations across the central visual field.

<<>>=
default_locmap <- getlocmap()
str(default_locmap, strict.width = "wrap")
@

Each \code{locmap} contains:
\begin{itemize}
  \item A coordinate table with $x$- and $y$-positions (in degrees),
  \item A vector \code{bs} indicating locations corresponding to the physiological blind spot.
\end{itemize}

Other patterns are available in the package, including the \code{10-2} grid.

<<>>=
ten_two_locmap <- locmaps$p10d2
str(ten_two_locmap, strict.width = "wrap")
@

If the blind spot is outside the tested region (e.g., in the 10-2 pattern) or is skipped (e.g., in the Octopus G1), \code{bs} is set to \code{numeric(0)}.

\subsection{Working with visual field data}

The package includes example data from a glaucoma patient (\code{vfpwgSunyiu24d2}), consisting of 42 visual field tests.

<<>>=
str(vfpwgSunyiu24d2[, 1:10], strict.width = "wrap")
@

Each row represents a monocular test. The eye is labeled using Latin abbreviations:

\begin{description}
\item [OD] oculus dexter -> right eye
\item [OS] oculus sinister -> left eye
\item [OU] oculus uterque -> both eyes
\end{description}

Initial columns store patient information and reliability metrics. Actual test values (sensitivities) begin at the column number stored in the \code{locini} variable. The function \code{getvfcols()} returns the numbers of the columns that contain the actual test values at the locations \code{l1} to \code{ln}, where 1..n corresponds to the rows of \code{coords} in the \code{locmaps}.

<<>>=
getlocini()
vfpwgSunyiu24d2[1, getvfcols()]
@

No columns should appear after these test values. Use \code{vffilter()} to select specific tests and \code{vfisvalid()} to verify data structure:

@@

The \code{vfread()} and \code{vfread()} functions can be used to read and write data in this format from a CSV file. However, these functions cannot be used to import native data from the perimeters. For this, the package includes functions to import data from the Octopus 900 and the Humphrey Field Analyzer.

\begin{itemize}
  \item \code{loadoctopus()} for Octopus 900 (Haag-Streit, Switzerland)
  \item \code{loadhfaxml()}, \code{loadhfadicom()}, \code{loadhfaxmlbatch()}, \code{loadhfadicombatch()} for the Humphrey Field Analyzer (Zeiss)
\end{itemize}

Detailed examples for exporting data from the Octopus 900 and using \code{loadoctopus()} can be found here: \url{https://huchzi.github.io/vf1/loadOctopus.html}.

\subsection{Normative data}

@@

To detect abnormal points, we compare observed sensitivities with an age-corrected mean normal field. These comparisons require a \emph{normative dataset} and location-wise \emph{age-linear models}.

The difference between observed and normal values is the \emph{total deviation (TD)}. In this package, negative values indicate lower-than-normal sensitivity, following the Humphrey Field Analyzer convention. For perimeters that reverse this definition (e.g., Octopus), users must invert the sign.

To reveal focal defects masked by generalized depression (e.g., due to cataract), a \emph{pattern deviation (PD)} is computed by subtracting a general height (GH) estimate from the TD. For 24-2, GH is defined as the 7th highest TD value.

The package includes three datasets with visual fields from normal observers:

\begin{itemize}
  \item SUNY-IU dataset (\code{vfctrsunyiu_24d2}) for 24-2 with Goldmann size III \citep{IMF:Mar2013b}
  \item Iowa PC26 (\code{vfctriowa_PC26}) and Peri (\code{vfctrsunyiu_10d2}) datasets for Goldmann size V \citep{IMF:Marin2018c}
\end{itemize}

<<>>=
str(
  vfctrSunyiu24d2, 
  strict.width = "wrap")
@

The normal values derived from these data can be found in \code{normvals}.

<<>>=
str(
  normvals$sunyiu_24d2, 
  max.level = 1, strict.width = "wrap")
@

To build the normative model for a custom dataset, use \code{nvgenerate()}.

@@

The normative object \code{nv} includes coefficients of age-linear models (\code{nv$agem$coeff}).:

<<>>=
head(nv$agem$coeff, 5)
@

It also contains a function that calculates normal sensitivity for a given age (\code{nv$agem$model()}), Thus, the expected field at age 50 is:

<<>>=
age <- 50
nv$agem$model(age)
@

Blind spot locations are returned as \code{NA}.

\subsection{Metric for individual locations}

@@

Total deviation is the difference between observed and expected sensitivity. Here, it is shown how it is calculated.

@@

To calculate the total deviation, use the function from the \code{nv} object:

@@

Global indices summarize the visual field by applying summary statistics on sensitivity, total deviation or pattern deviation. These indices are calculated using \code{getgl()}. The resulting \code{data.frame} also contains the patient's age, eye, and date of the test.

<<>>=
gl <- getgl(example2)
str(gl, max.level = 1, strict.width = "wrap")
gl[, -(1:getlocini() - 1)]
@

\begin{description}
  \item [msens]: Mean sensitivity (MS)
  \item [ssens]: SD of sensitivity
  \item [tmd]: Mean total deviation (MD)
  \item [tsd]: SD of total deviation
  \item [pmd]: Mean pattern deviation
  \item [psd]: SD of pattern deviation
  \item [gh]: General height
  \item [vfi]: Visual field index (VFI) \citep{IMF:Ben2008}
\end{description}

Associated probabilities are calculated with \code{getglp()}:

<<>>=
getglp(gl)[, -(1:getlocini() - 1)]
@

@@

Standard displays include color-coded probability maps, graphical representations of the blind spot, and border enhancements that encode diagnostic information. For example, \code{vfplot} with the argument 	\code{"td"} generates plots that require normative data on differential light sensitivity and the probabilities of deviations from these norms. This information is included for the Humphrey Field Analyzer patterns but must be provided separately for other perimeters.

@@

It is important to specify the type of data represented in the visual field tables. The default is set to \emph{total deviation} (	exttt{td}).

@@

Progression analysis methods fall into two broad categories: \emph{event-based} and \emph{trend-based} approaches. These may further be classified by their reliance on population-based normative data, mathematical models of progression, or model-free methods.

@@

\subsubsection{Plotting progression}

@@

\subsubsection{Statistical analysis}

Statistical progression analysis may be performed via:
\begin{itemize}
\item \emph{Global linear regression} (e.g., on MD)
\item \emph{Pointwise linear regression} (PLR)
\end{itemize}

The \code{glr()} function estimates a regression model for global indices. For the example series, the patient loses \code{getgl(example3)} dB/year, with $p < getgl(example3)$.

%\code{r round(abs(glr(getgl(example3))$sl), 2)}
%\code{r format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)}

Pointwise linear regression is performed using \code{plr()}, resulting in one slope and $p$-value per location:

<<>>=
prog1 <- plr(example3)
str(prog1, max.level = 1, strict.width = "wrap")
print(sum(prog1$pval[-getlocmap()$bs] < 0.05))
@

@@

PLR-based criteria (e.g., 3 or more progressing points) are clinically useful but lack explicit control for specificity. \emph{Permutation of Pointwise Linear Regression (PoPLR)} addresses this by combining the 52 $p$-values into a single $S$-statistic (per Fisher's method) and comparing it to a null distribution from permuted series \citep{IMF:Ole2012,IMF:Marin2021}.

<<>>=
prog2 <- poplr(example3)
str(prog2, max.level = 1, strict.width = "wrap")
@

@@

\subsection{Automated reports}

@@

The ganglion cell bodies stimulated by a visual field location are displaced from the photoreceptor input region, especially in the central field. This displacement is modeled in \pkg{visualFields} based on Montesano et al. \citep{IMF:Montesano2020}.

@@

Nerve fiber paths from each location to the optic disc can also be modeled, using Jansonius' framework \citep{IMF:Jan2009, IMF:Jan2012}.

<<>>=
psi23 <- loc2psi(vf2gc(locmaps$p24d2$coord[23,]))
ang23 <- psi2oct(psi23)

ang23 - 360
@

@@

Perimetry conventions vary across manufacturers. The package includes normative data for Humphrey 24-2 (Goldmann III) and Iowa full-field perimetry with Goldmann V \citep{IMF:Marin2018c}. The \code{loadoctopus} function provides partial support for Octopus G1 fields, but normative support is incomplete.

Users may provide custom normative datasets using \code{nvgenerate()} and \code{setnv()}.

@@

Each analysis requires environment variables:
\begin{itemize}
\item \code{nv}: normative values
\item \code{locmap}: test grid
\item \code{gpar}: graphical parameters
\item \code{locini}: first sensitivity column
\end{itemize}

<<>>=
names(locmaps)
names(normvals)
names(gpars)
@

@@

Custom test structures must:
\begin{itemize}
\item Start with columns: \code{id}, \code{eye}, \code{date}, \code{time}
\item Include \code{age}, optionally \code{fpr}, \code{fnr}, \code{fl}
\item Have sensitivity columns named \code{l1}--\code{lk}
\end{itemize}

@@

If normative data exist, use \code{vfread()}, \code{nvgenerate()}, and \code{setnv()} to enable deviation computations.
